---
title: "Exploring NMF and other views of high-dimensional multivariate data"
shorttitle: "Dimension reduction concepts"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Exploring NMF and other views of high-dimensional multivariate data}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

To improve my
understanding of the role of the HALS algorithm for NMF analysis
of genomic data, I work with the ALL microarray dataset.

```
     Sabina Chiaretti, Xiaochun Li, Robert Gentleman, Antonella Vitale,
     Marco Vignetti, Franco Mandelli, Jerome Ritz, and Robin Foa Gene
     expression profile of adult T-cell acute lymphocytic leukemia
     identifies distinct subsets of patients with different response to
     therapy and survival. Blood, 1 April 2004, Vol. 103, No. 7.
```

```{r getlibs,message=FALSE,results="hide"}
# to avoid conflicting symbol warnings
suppressWarnings({
library(ALL)
library(matrixStats)
library(hgu95av2.db)
library(SummarizedExperiment)
library(randomForest)
})
```

# Data acquisition and setup

## Filtering and feature annotation

I reduce the dataset down to two patient types:
those with BCR/ABL fusion and those negative
for any known cytogenetic feature used
in prognosis at the time of the study.
```{r lk1,message=FALSE}
library(ALL)
data(ALL)
fuvn = ALL[, which(ALL$mol.biol %in% c("BCR/ABL", "NEG"))]
fuvn$mol.biol = factor(fuvn$mol.biol)
```

I filter the features down to those
in the upper quantile of between-patient
variation.

```{r dofilt,message=FALSE}
library(matrixStats)
ss = rowSds(exprs(fuvn))
#> sum(ss>.5)
#[1] 3182
ffuvn = fuvn[which(ss>.5),]
```

Transform the data representation to SummarizedExperiment
and rename the features with HUGO symbols.
```{r rena, message=FALSE}
library(hgu95av2.db)
mapas = mapIds(hgu95av2.db, keys=rownames(ffuvn), column="SYMBOL", keytype="PROBEID")
# as = affy to symbol
library(SummarizedExperiment)
ffuvnse = as(ffuvn, "SummarizedExperiment")
ffuvnse
bad = which(is.na(mapas))
pbad = names(mapas)[bad]
exclprobes = intersect(pbad, rownames(ffuvnse))
badinds = which(rownames(ffuvnse) %in% exclprobes)
ffuvnse = ffuvnse[-badinds,]
rowData(ffuvnse) = DataFrame(probeid=rownames(ffuvnse), sym=mapas[rownames(ffuvnse)])
rownames(ffuvnse) = rowData(ffuvnse)$sym
ffuvnse
```

# Isolation of informative features using random forests

```{r getrf, message=FALSE, cache=TRUE}
library(randomForest)
set.seed(1234) # make reproducible
ldf = data.frame(molb=ffuvnse$mol.biol, t(assay(ffuvnse)))
rf1 = randomForest(molb~., data=ldf, importance=TRUE)
rf1
```
Even though the overall OOB misclassification estimate is attractive,
the confusion matrix is not very appealing.

```{r lkvari}
varImpPlot(rf1)
```

Let's take the top 15 features for multivariate analysis.

```{r gettop}
im = rf1$importance
top = rownames(im[order(im[,3], decreasing=TRUE),])[1:15]
top
```

# Multivariate exploration 1: PCA

Basic computation:
```{r dopc1}
pp = prcomp(data.matrix(ldf[,top]))
```

Overview:
```{r lkpcs}
pairs(pp$x[,1:4], col=factor(ldf$molb), pch=19, cex=.4)
```

A simple biplot to show some details
of the projection (first two components)
```{r dopc1b}
ggbiplot(pp) + geom_point(aes(colour=ldf$molb))
```
Try 2 and 3:
```{r dopc2}
ggbiplot(pp, c(2,3)) + geom_point(aes(colour=ldf$molb))
```

While PC1 seems to provide information to discriminate
BCR/ABL fusion from other ALL negative for prognostic cytogenetic
features, PC2 and PC3 may be helping to isolate genes
that discriminate different transcriptional
patterns (and possibly different clinical
sequelae) within BCR/ABL fusion.  To confirm this,
one could work with samples from the TARGET study using
Bioconductor's [GenomicDataCommons](https://bioconductor.org/packages/release/bioc/html/GenomicDataCommons.html)
package.
